plyranges: A grammar of genomic data transformation

### 2018-09-18 paper.Rmd
reorder of design, added paragraph about memory usage;  extended note about summarized experiment

## Genomic Relation Algebra

@@

For merging based on the hits, we have three modes: _inner_,
_intersect_ and _left_. The _inner_ overlap join is similar to the
conventional inner join in that there is a row in the result for every
match. A major difference is that the matching is not by identity, so
we have to choose one of the ranges from each pair. We always choose
the left range. The _intersect_ join uses the intersection instead of
the left range. Finally, the overlap _left_ join is akin to left outer
join in Codd's relational algebra: it performs an overlap inner join
but also returns all query ranges that are not hit by the subject.

@@

Like `dplyr`, `plyranges` verbs are functional: they are free of side
effects and return their result. This enables chaining of verbs
through syntax like the forward pipe operator (`%>%`, read as "then")
of the `magrittr` package \cite{R-magrittr}. This has syntax has a direct
cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via sub-replacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to
so-called fluent syntax.

A function is cohesive if it performs a singular task without
producing any side-effects. Singular tasks can always be broken 
down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the `GenomicRanges` package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, `plyranges` increases cohesion and
decouples the anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names
as the primary means of expression, instead of a more complex mixture
of function and argument names. This results in the
user being able to conceptualize the `plyranges` DSL as a flat catalog of 
functions, without having to descend further into documentation to understand
a function's arguments. A flat function catalog also enhances API
discoverability, particularly through auto-completion in IDEs. One
downside of pushing cohesion to this extreme is that function calls
become coupled, and care is necessary to treat them as a group when
modifying code.

@@

@@

We aim to continue developing the `plyranges` package and to extend it
for use with more complex data structures, such as the
SummarizedExperiment class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. The
grammar and design of the `plryanges` DSL are naturally extensible to
SummarizedExperiment.

>>>>

## Genomic Relational Algebra

@@

For merging based on the hits, we have three modes: _inner_,
_intersect_ and _left_. The _inner_ overlap join is similar to the
conventional inner join in that there is a row in the result for every
match. A major difference is that the matching is not by identity, so
we have to choose one of the ranges from each pair. We always choose
the left range. The _intersect_ join uses the intersection instead of
the left range. Finally, the overlap _left_ join is akin to left outer
join in Codd's relational algebra: it performs an overlap inner join
but also returns all _x_ ranges that are not hit by the _y_ ranges.

@@

To further enable cognitive consistency, `plyranges` functions are cohesive. 
A function is defined to be cohesive if it performs a singular task without
producing any side-effects. Singular tasks can always be broken 
down further at lower levels of abstraction.
For example, to resize a range, the user needs to specify which
position (start, end, midpoint) should be invariant over the
transformation. The `resize()` function from the `GenomicRanges` package
has a `fix` argument that sets the anchor, so calling `resize()`
coalesces anchoring and width modification. The coupling at the
function call level is justified since the effect of setting the width
depends on the anchor. However, `plyranges` increases cohesion and
decouples the anchoring into its own function call.

Increasing cohesion simplifies the interface to each operation, makes
the meaning of arguments more intuitive, and relies on function names
as the primary means of expression, instead of a more complex mixture
of function and argument names. This results in the
user being able to conceptualize the `plyranges` DSL as a flat catalog of 
functions, without having to descend further into documentation to understand
a function's arguments. A flat function catalog also enhances API
discoverability, particularly through auto-completion in 
integrated developer environments (IDEs). One downside of pushing cohesion to 
this extreme is that function calls become coupled, and care is 
necessary to treat them as a group when modifying code.

Like `dplyr`, `plyranges` verbs are functional: they are free of side
effects and are generally endomorphic, meaning that when the input is a GRanges
object they return a GRanges object. This enables chaining of verbs
through syntax like the forward pipe operator from the `magrittr` package. 
This syntax has a direct cognitive mapping to natural language and the intuitive notion of
pipelines. The low-level object-oriented APIs of Bioconductor tend to
manipulate data via sub-replacement functions, like `start(gr) <-
x`. These ultimately produce the side effect of replacing a symbol
mapping in the current environment and thus are not amenable to so-called 
fluent syntax.

@@

The implementation of `plyranges` is built on top of 
Bioconductor infrastructure, meaning most functions are constructed
by composing generic functions from core Bioconductor packages. As a result, any 
Bioconductor packages that uses data structures that inherit from GRanges will
be able to use `plyranges`  for free. Another consequence of building on top of
Bioconductor generics is that the speed and memory usage of `plyranges` 
functions are similar to the highly optimised methods implemented in Bioconductor
for GRanges objects.

@@

We aim to continue developing the `plyranges` package and to extend it
for use with more complex data structures, such as the
SummarizedExperiment class, the core Bioconductor data structure for
representing experimental results (e.g., counts) from multiple sample
experiments in conjunction with feature and sample metadata. Although,
the SummarizedExperiment is not strictly tidy, it does consist of 
three tidy data structures that are related by feature and sample identifiers. 
Thererfore, the grammar and design of the `plryanges` DSL is naturally 
extensible to the SummarizedExperiment.
